**Reference-based OTU clustering for machine learning classification**

### 2022-06-20 submission/manuscript.Rmd
revised ms

Other Title Options:\
A solution to the problem of inconsistent OTU clustering for machine
learning classification

Optimized reference-based OTU clustering for machine learning
classification

@@

Running title: INSERT RUNNING TITLE HERE

@@

## Abstract (250 word max)

Machine learning classification of disease based on the gut microbiome
often relies on operational taxonomic units (OTUs) to quantify microbial
composition. The standard methodology for quantifying taxonomic
composition is clustering 16S rRNA amplicon sequences *de novo* into
OTUs and using their abundances to train a classification model.
However, OTU clusters depend on the sequences in the data set and
therefore can change if new data is added. This inconsistency
complicates machine learning classification because in order to use the
model to classify additional samples, the OTUs have to be re-clustered
to include the new sequences and the model retrained with the new OTU
clusters. A new algorithm in mothur, called OptiFit, addresses this
issue by clustering new sequences into the existing OTU clusters. While
OptiFit is proven to produce high quality OTU clusters, it is unclear
whether this method for clustering new sequence data into existing OTUs
will have any impact on machine learning classification using the OTU
abundances. To evaluate the potential for use in machine learning
classification, we used OptiFit to cluster additional data into existing
OTU clusters and quantified model performance in classifying a data set
containing samples from controls and people with advanced adenoma and
carcinoma (SRN). We compared the model performance using OptiFit to the
standard procedure of clustering all the data together. We found that
OptiFit can be used to fit new sequence data to existing OTUs and
perform equally in classifying SRNs. Moving forward, when OTUs are used
in machine learning classification problems, OptiFit can be used to
avoid the need to re-cluster sequences into OTUs when classifying
additional samples that were not part of the data set used to build the
model.

## Importance (150 word max)

There are instances when OTU abundance data is optimal for machine
learning prediction of disease based on microbiome composition. The
current methodology for OTU clustering complicates machine learning
since *de novo* OTU clusters can change depending on the data in the
data set. Attempting to classify additional samples that were not part
of the original model build requires re-clustering OTUs and re-training
the model. OptiFit provides an elegant solution to the problem of OTU
inconsistency by enabling new data to be fit to the existing OTUs while
maintaining machine learning classification performance. The use of
OptiFit will enable a single model based on OTU abundance data to be
used in classifying additional samples that were not part of the
original OTU clustering.

@@

Gut microbiome community composition is useful as a resource for machine
learning prediction of diseases, including colorectal cancer
[@baxter2016; @zackular2014]. Amplicon sequencing of the 16S rRNA gene
is a reliable tool for assessing the taxonomic composition of microbial
communities. Analysis of 16S rRNA amplicon sequence (16S) data generally
relies on clustering of sequences based on similarity into operational
taxonomic units (OTUs). The process of OTU clustering can either be
reference-based or *de novo*. The quality of OTUs generated with
reference-based clustering is dependent on the quality of the reference
used, while OTUs generated with *de novo* clustering are optimized to
the data set @westcott2015. While *de novo* clustering can produce
high-quality OTU clusters where sequences are accurately grouped based
on similarity thresholds, the resulting OTU clusters depend on the data
in the data set and the addition of new data could change the overall
OTU clusters. The inconsistent nature of OTU clustering complicates
deployment of machine learning models since integration of additional
data requires re-clustering all the data and re-training of the model.
The ability to integrate new data into an existing model without
re-clustering and re-training could allow for deployment of a single
model that new data can be continually added to. Recently Sovacool *et
al* introduced OptiFit: a method for fitting new data into existing OTU
clusters @sovacool2022. While OptiFit is proven to effectively fit new
sequence data to existing OTU clusters, it is unknown if the use of
OptiFit will have an impact on machine learning classification. Here we
test the ability of OptiFit to cluster new sequence data into existing
OTU clusters for the purpose of machine learning classification of
disease based on gut microbiome composition.

When using OTU abundances for machine learning classification, the
current methodology is to *de novo* cluster all of the sequence data
into OTUs with the OptiClust algorithm in mothur @westcott2017. The
resulting abundance data is then split into training and testing sets,
where the training set is used to tune hyperparameters and ultimately
train the model. The testing set is then classified with the model and
the performance of the model can be quantified (Figure 1A). However,
with this methodology we would have to re-generate the OTU clusters and
re-train the model if we wanted to classify additional samples. The
OptiFit algorithm @sovacool2022 addresses this problem by enabling new
sequences to be clustered into existing OTUs. The OptiFit workflow is
similar to the OptiClust workflow where the data is clustered into OTUs
and used to tune hyperparameters and ultimately train the model. Then,
we can use OptiFit to fit sequence data of samples not part of the
original data set into the existing OTUs and use the same model to
classify the samples (Figure 1B). To test how the model performance
compares between these two methodologies, we used a publicly available
data set of 16S sequences from stool samples of healthy subjects as well
as subjects with SRN consisting of advanced adenoma and carcinoma
@baxter2016. The data set was randomly split into an 80% train set and
20% test set. For the standard OptiClust workflow, the training and test
sets were *de novo* clustered together into OTUs then the resulting
abundance table was split into the training and testing set. For the
OptiFit workflow, the train set was clustered *de novo* into OTUs and
the remaining test set was fit to the OTU clusters using the OptiFit
algorithm. For both workflows, the abundance table of the train set was
used to tune hyperparameters and train a random forest model to classify
SRN. The test set was classified as either control or SRN using the
trained models.To account for variation depending on the split of the
data, the data set was randomly split 100 times and the process repeated
for each of the 100 data splits. By comparing the model performance of
classifying the samples in the test data set between the OptiFit and
OptiClust algorithms, we can quantify the impact of using OptiFit on
model classification performance.

@@

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score is quantified by examining all pairs of sequences and assessing
whether they belong together in an OTU or not based on sequence
similarity @westcott2017. MCC scores range between negative one and one.
A score of negative one means none of the sequences in an OTU are within
the similarity threshold and any sequences within the similarity
threshold are not in an OTU together. An MCC score of zero essentially
means the sequences are randomly clustered. An MCC score of 1 means all
sequences in an OTU are within the similarity threshold and all sequence
pairs within the similarity threshold are in the same OTU. To ensure
that OptiFit is appropriately integrating new sequence data into the
existing OTUs, we would expect that the MCC scores produced by the
OptiClust and OptiFit workflows are similar. Since the data is only
clustered once in the OptiClust workflow there is only one MCC score
while the OptiFit workflow produces an MCC score for the OTU clusters
from each data split. Overall the MCC scores were similar between
OptiClust (MCC = `r round(opticlust_mcc,digits=3)`) and OptiFit (average
MCC = `r round(optifit_avg_mcc,digits=3)`) indicating that OptiFit
performs as well as OptiClust when integrating new sequences into the
existing OTUs.

@@

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we next examined the model performance for
classifying samples in the test data set as control or SRN. To quantify
model performance we used the taxonomic abundances of the training data
from the OptiClust and OptiFit workflows to train a model to predict
SRNs. Using the predicted and actual diagnosis classification, we
calculated the area under the receiver operating characteristic curve
(AUROC) for each data split to quantify model performance. During
cross-validation (CV) training, the model performance was equivalent
between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
OptiClust mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
OptiFit mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`,
Figure 2A). The trained model was then deployed to classify the samples
of the test data as control or SRN. The performance on the test data was
equivalent between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
OptiClust mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
OptiFit mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
Figure 2B,C) indicating that new data can be fit to existing OTU
clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification of disease based
on microbiome abundance. A potential problem to using OptiFit for
machine learning prediction is that any sequences in the new data that
do not map to the existing OTU clusters will be discarded resulting in a
possible loss of information. However, we demonstrated that OptiFit can
be used to fit new sequence data into existing OTU clusters and perform
equally in predicting SRN compared to clustering all of the sequence
data together. The ability to integrate new data into existing OTUs
enables the deployment of a single machine learning model based on
microbiome composition that new data can be classified with. These
results are based on a single data set and disease. Further analysis is
needed to determine the data size necessary to build a robust model
capable of classifying diverse data sets. A robust machine learning
model could be implemented as part of a non-invasive and low-cost aid in
diagnosing SRN.

@@

***Data Set.*** Raw 16S rRNA amplicon sequence data isolated from human
stool samples was downloaded from NCBI Sequence Read Archive (accession
no. SRP062005) @baxter2016.; @edgar2011 This data set contains stool
samples from a total of N subjects, however after preprocessing to
screen for sequence quality and subsample to 10,000 reads per sample,
490 samples remained. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as "normal" while samples from subjects identified as
advanced adenoma or carcinoma were categorized as "screen relevant
neoplasia" (SRN). The resulting data set consisted of 261 normal samples
and 229 SRN samples.

@@

***Data Processing.*** The full dataset was pre-processed with mothur
(v1.45) @schloss2009 using the SILVA reference database (v132)
@quast2013 to join forward and reverse reads, merge duplicate reads,
align to the reference, pre-cluster, remove chimeras with UCHIME
@edgar2011, assign taxonomy, and remove non-bacterial reads following
the Schloss Lab MiSeq standard operating procedure described on the
mothur website (<https://mothur.org/wiki/miseq_sop/>). 100 splits of the
490 samples were generated where 80% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20% (98 samples)
were assigned to the test set. Using 100 splits of the data accounts for
the variation that may be observed depending on the samples that are in
the training or test sets. Each samples was in the training set an
average of `r n_test_train %>% pull(avg_train)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
the test set an average of `r n_test_train %>% pull(avg_test)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).

The data was processed through two workflows. First, the standard
workflow using the OptiClust algorithm @westcott2017. In this pathway,
all of the data was clustered together with OptiClust to generate OTUs
and the resulting abundance tables were split into the training and
testing sets. In the second workflow, the pre-processed data was split
into the training and testing sets. The training set was clustered into
OTUs, then the test set was fit to the OTUs of the training set using
the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with
method open so that any sequences that didn't map to the existing OTU
clusters would form new OTUs. Any OTUs that were not in the training set
were removed prior to machine learning. For both pathways, the shared
files were sub-sampled to 10,000 reads per sample.

***Machine Learning.*** Machine learning using Random Forest was
conducted with the R package mikrompl (v XXXX) @topçuoglu2021 to predict
the diagnosis (SRN or normal) for the samples in the test set for each
data split. The training set was preprocessed to normalize values
(scale/center), collapse correlated features, and remove features with
zero-variance. The preprocessing from the training set was then applied
to the test set. P values comparing model performance were calculated as
previously described {}. The averaged ROC curves were plotted by taking
the average and standard deviation of the sensitivity at each
specificity value.

@@

(funding)

>>>>

@@

Running title: Reference-based OTU clustering for ML classification

@@

## Abstract

Machine learning classification of disease based on the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The standard
approach to clustering sequences into OTUs leverages the similarity of
the sequences to each other rather than to a reference database. The
abundance of each OTU is used to train a classification model. However,
OTU assignments depend on the sequences in the data set and therefore
can change if new data are added. This lack of stability complicates
classification because in order to use the model to classify additional
samples, the OTUs have to be reclustered to include the new sequences
and the model retrained with the new OTU clusters. A new reference-based
clustering algorithm, called OptiFit, addresses this issue by fitting
new sequences into existing OTUs. While OptiFit is proven to produce
high quality OTU clusters, it is unclear whether this method for fitting
new sequence data into existing OTUs will impact the performance of
classification models. We used OptiFit to cluster additional data into
existing OTU clusters and quantified model performance in classifying a
data set containing samples from patients with and without colonic
screen relevant neoplasias (SRN). We compared the performance of this
model to the standard procedure of clustering all the data together. We
found that both approaches performed equally well in classifying SRNs.
Moving forward, when OTUs are used in classification problems, OptiFit
can be used to avoid the need to retrain models using reclustered
sequences when classifying new samples.

## Importance

There is great potential for using microbiome data to non-invasively
diagnose people. One of the challenges with using classification models
generated using the relative abundance of operational taxonomic unit
(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
their similarity to other sequences in the dataset. If data are
generated from new patients seeking a diagnosis, then it would be
necessary to reassign sequences to OTUs and retrain the classification
model. Yet there is a desire to have a single, validated model that can
be deployed. To overcome this obstacle, we applied the OptiFit
clustering algorithm which fits new sequence data to existing OTUs
allowing the reuse of a consistent model. A random forest machine
learning model deployed using OptiFit performed as well as the
traditional reassignment and retraining approach. This result indicates
there is potential for developing machine learning classification models
based on OTU relative abundance data.

@@

Gut community composition is useful as a resource for machine learning
classification of diseases, including colorectal cancer [@baxter2016;
@zackular2014]. Amplicon sequencing of the 16S rRNA gene is a reliable
tool for assessing the taxonomic composition of microbial communities,
which is the input to these models. Analysis of 16S rRNA gene sequence
data generally relies on clustering of sequences based on similarity
into operational taxonomic units (OTUs). The process of OTU clustering
can either be reference-based or *de novo*. The quality of OTUs
generated with reference-based clustering is generally poor compared to
those generated with *de novo* clustering @westcott2015. While *de novo*
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the data in the data set and the addition of new data
could change the overall OTU clusters. The unstable nature of OTU
clustering complicates deployment of machine learning models since
integration of additional data requires reclustering all the data and
retraining of the model. The ability to integrate new data into a
validated model without reclustering and retraining could allow for
deployment of a single model that new data can be continually added to.
Recently Sovacool *et al* introduced OptiFit: a method for fitting new
sequence data into existing OTUs @sovacool2022. While OptiFit is proven
to effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification.
Here we tested the ability of OptiFit to cluster new sequence data into
existing OTU clusters for the purpose of classification of disease based
on gut microbiome composition.

We compared two approaches, one using all of the data to generate OTU
clusters and the other generating OTU clusters with a portion of the
data and then fitting the remaining sequence data to the existing OTUs
using OptiFit. In the first approach, all of the 16S rRNA sequence data
was *de novo* clustered into OTUs with the OptiClust algorithm in mothur
@westcott2017. The resulting abundance data was then split into training
and testing sets, where the training set was used to tune
hyperparameters and ultimately train the model. The testing set was then
classified with the model and the performance of the model was
quantified (Figure 1A). However, with this methodology we would have to
regenerate the OTU clusters and retrain the model if we wanted to
classify additional samples. The OptiFit algorithm @sovacool2022
addresses this problem by enabling new sequences to be clustered into
existing OTUs. The OptiFit workflow is similar to the OptiClust workflow
where the data was clustered into OTUs and used to tune hyperparameters
and ultimately train the model. Then, we used OptiFit to fit sequence
data of samples not part of the original data set into the existing OTUs
and used the same model to classify the samples (Figure 1B). To test how
the model performance compares between these two methodologies, we used
a publicly available data set of 16S rRNA gene sequences from stool
samples of healthy subjects as well as subjects with SRN consisting of
advanced adenoma and carcinoma @baxter2016. The data set was randomly
split into an 80% train set and 20% test set. For the standard OptiClust
workflow, the training and test sets were *de novo* clustered together
into OTUs then the resulting abundance table was split into the training
and testing set. For the OptiFit workflow, the train set was clustered
*de novo* into OTUs and the remaining test set was fit to the OTU
clusters using the OptiFit algorithm. For both workflows, the abundance
table of the train set was used to tune hyperparameters and train a
random forest model to classify SRN. The test set was classified as
either control or SRN using the trained models.To account for variation
depending on the split of the data, the data set was randomly split 100
times and the process repeated for each of the 100 data splits. By
comparing the model performance of classifying the samples in the test
data set between the OptiFit and OptiClust algorithms, we quantified the
impact of using OptiFit on model classification performance.

@@

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score was quantified by examining all pairs of sequences and assessing
whether they belonged together in an OTU based on their similarity
@westcott2017. MCC scores range between negative one and one. A score of
negative one means none of the sequences in an OTU are within the
similarity threshold and any sequences within the similarity threshold
are not in an OTU together. An MCC score of zero essentially means the
sequences are randomly clustered. An MCC score of 1 means all sequences
in an OTU are within the similarity threshold and all sequence pairs
within the similarity threshold are in the same OTU. To ensure that
OptiFit is appropriately integrating new sequence data into the existing
OTUs, we expected the MCC scores produced by the OptiClust and OptiFit
workflows to be similar. Since the data was only clustered once in the
OptiClust workflow there was only one MCC score while the OptiFit
workflow produced an MCC score for the OTU clusters from each data
split. Overall the MCC scores were similar between OptiClust (MCC =
`r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
`r round(optifit_avg_mcc,digits=3)`). This indicated that OptiFit
performed as well as OptiClust when integrating new sequences into the
existing OTUs.

@@

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we examined the model performance for classifying
samples in the held out test data set. To quantify model performance we
used the OTU relative abundances from the training data from the
OptiClust and OptiFit workflows to train a model to predict SRNs. Using
the predicted and actual diagnosis classification, we calculated the
area under the receiver operating characteristic curve (AUROC) for each
data split to quantify model performance. During cross-validation (CV)
training, the model performance was equivalent between the two
algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
OptiClust mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
OptiFit mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`,
Figure 2A). The trained model was then deployed to classify the samples
of the test data as control or SRN. The performance on the test data was
equivalent between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
OptiClust mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
OptiFit mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
Figure 2B,C) indicating that new data could be fit to existing OTU
clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences in the new data that do not map to the existing OTU clusters
will be discarded resulting in a possible loss of information. However,
we demonstrated that OptiFit can be used to fit new sequence data into
existing OTU clusters and perform equally well in predicting SRN
compared to clustering all of the sequence data together. The ability to
integrate data from new samples into existing OTUs enables the
deployment of a single machine learning model. These results are based
on a single data set and disease. Further analysis is needed to
determine the number of samples that are necessary to build a robust
model capable of classifying diverse samples. A robust machine learning
model could be implemented as part of a non-invasive and low-cost aid in
diagnosing SRN and other diseases.

@@

***Data Set.*** Raw 16S rRNA gene sequence data isolated from human
stool samples was downloaded from NCBI Sequence Read Archive (accession
no. SRP062005) [@edgar2011; @baxter2016]. This data set contains stool
samples from a total of 490 subjects. For this analysis, samples from
subjects identified in the metadata as normal, high risk normal, or
adenoma were categorized as "normal" while samples from subjects
identified as advanced adenoma or carcinoma were categorized as "screen
relevant neoplasia" (SRN). The resulting data set consisted of 261
normal samples and 229 SRN samples.

@@

***Data Processing.*** The full dataset was preprocessed with mothur
(v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
reads, align to the SILVA reference database (v132) @quast2013,
precluster, remove chimeras with UCHIME @edgar2011, assign taxonomy, and
remove non-bacterial reads following the Schloss Lab MiSeq standard
operating procedure described on the mothur website
(<https://mothur.org/wiki/miseq_sop/>). 100 splits of the 490 samples
were generated where 80% of the samples (392 samples) were randomly
assigned to the training set and the remaining 20% (98 samples) were
assigned to the test set. Using 100 splits of the data accounts for the
variation that may be observed depending on the samples that are in the
training or test sets. Each sample was in the training set an average of
`r n_test_train %>% pull(avg_train)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
the test set an average of `r n_test_train %>% pull(avg_test)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).

The data was processed through two workflows. First, the standard
workflow using the OptiClust algorithm @westcott2017. In this pathway,
all of the data was clustered together with OptiClust to generate OTUs
and the resulting abundance tables were split into the training and
testing sets. In the second workflow, the preprocessed data was split
into the training and testing sets. The training set was clustered into
OTUs, then the test set was fit to the OTUs of the training set using
the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with
method open so that any sequences that did not map to the existing OTU
clusters would form new OTUs. For both pathways, the shared files were
sub-sampled to 10,000 reads per sample.

***Machine Learning.*** Machine learning using Random Forest was
conducted with the R package mikrompl (v 1.2.0) @topçuoglu2021 to
predict the diagnosis (SRN or normal) for the samples in the test set
for each data split. The training set was preprocessed to normalize OTU
counts (scale/center), collapse correlated OTUs, and remove OTUs with
zero-variance. The preprocessing from the training set was then applied
to the test set. Any OTUs in the test set that were not in the training
set were removed. P values comparing model performance were calculated
as previously described @topçuoglu2020. The averaged ROC curves were
plotted by taking the average and standard deviation of the sensitivity
at each specificity value.

@@

This work was supported through a grant from the NIH (R01CA215574).
